%
% introduction.tex
%
% Copyright (C) 2020 by SpaceLab.
%
% DOCUMENTATION-TEMPLATE
%
% This work is licensed under the Creative Commons Attribution-ShareAlike 4.0
% International License. To view a copy of this license,
% visit http://creativecommons.org/licenses/by-sa/4.0/.
%

%
% \brief Introduction chapter.
%
% \author Gabriel Mariano Marcelino <gabriel.mm8@gmail.com>
%
% \institution Universidade Federal de Santa Catarina (UFSC)
%
% \version 0.1.0
%
% \date 2020/07/16
%

\chapter{Work Plan} \label{ch:work-plan}

\paragraph{}
\indent 
   To achieve the proposed objective, the following tasks must be concluded:
   
   \begin{enumerate}
   	\item Develop a model for the mechanical damping effect of the hysteresis loss.
   	\item Choose a simplified model for the geomagnetic field.
   	\item Choose a simplified model for the orbital dynamics.
   	\item Couple Euler's equations of motion for rigid bodies with the magnetic phenomena and with the orbit propagation model.
   	\item Define a post-processing strategy to analyze the effects of the attitude control on the communication links.
   	\item Write a report regarding the system performance evaluated with the simplified models.
   	\item Refine the geomagnetic field model.
   	\item Refine the orbit propagation model.
   	\item Refine the communication link analysis strategy.
   	\item Write a report regarding the system performance evaluated with the refined models.
   	\item Run a parametric analysis to search for better configurations of the attitude control system.
   	\item Write a report regarding the results of the parametric analysis and providing a conclusion about the necessity of a redesign.
   \end{enumerate}